---
title: "2016 Twitter"
output: html_document
---
Twitter is a popular social media network in which users can send and receive short 140-character messages ("tweets") on any topic they wish. We have collected 5000 tweets from Twitter on 27 June 2016, searching on the word "love." 


We will show how regular expressions and the package `stringr` can be used to analyze the text in this data set.


```{r}
library(stringr)
```


We will import the text file **love.txt** using the `scan` command. We specify that each line is a character string using the argument `what =""`


```{r}
love <- scan("data/love.txt", what = "")
head(love)

```


How many of these tweets mention the word "heart"?


```{r}
out <- str_detect(love, "heart")
head(out)
head(love[out])
sum(out)  
```

Thus, 114 tweets contain "heart".
But did we detect all tweets that had the word "heart?" What if the tweeter writes "Heart" or HEART" or uses some other combination of upper and lower case letters? 


```{r}
out2 <- str_detect(love, "[Hh][Ee][Aa][Rr][Tt]")
sum(out2)
```

So there were only 6 additional tweets that had some capitalization in some of the letters in the word "heart".


To specify an "or", use the vertical bar. For instance, to find the number of tweets that mentions either "boy" or "girl"


```{r}
out3 <- str_detect(love, "boy|girl")
sum(out3)
```



The word "you" comes up frequently in tweets, often being mentioned more than once in a given tweet. Suppose we wish to count all occurrences of "you".


```{r}
out4 <- str_locate_all(love, "[Yy][Oo][Uu]")
head(out4)
head(unlist(out4))
```

The R object `out4` is a list. The `unlist` command creates a vector from the list. For instance, in the third tweet, "you" occurs in positions 62-64, 80-82, and 98-100. When you apply `unlist` to `out4`, these values appear in the vector as 62 80 90 64 82 100 (start positions first, followed by end positions).

Now, in `unlist(out4)`, the start and ending position of every occurrence of "you" is given. To find the total number of occurrences of "you", we must divide by 2.

```{r}
length(unlist(out4))/2
```

So of the 5000 tweets in this file, the word "you" occurred 3229 times.


Now, of the tweets in the file, how many were retweets? We need to count the number of tweets that start out with the letters "RT". 


```{r}
out5 <- str_detect(love, "^RT")
sum(out5)
```


Suppose we wish to count the number of times hashtags are used in the tweets. We need to match the pattern "#" followed by some characters. In particular, the hashtag character "#" cannot be followed by a space. The expression "\\S" indicates 'not a space' while a "+" following it will specify one or more instances of "not a space":

```{r}
out6 <- str_locate_all(love, "#\\S+")
length(unlist(out6))/2
```


Now, many tweets contain web addresses. The URL's start with "http" or "https" followed immediately by "://" and then more characters. So, after "http" we need to indicate either 0 or 1 occurrence of the "s". This is done using "s?" After the double forward slashes, we match zero or more occurrences of any character.


```{r}
outURLS <- str_extract_all(love, "http(s?)(://).*")
head(outURLS)    #a list
head(unlist(outURLS))
```


Many users include special characters in their tweets such as emojis, foreign symbols, etc. However, when these tweets are exported to a text file, these symbols are replaced by their unicode number, a universal standard for encoding special characters. For instance, take a look at the 12th and 195th tweets in this data file:


```{r}
love[c(12, 195)]   
```


`<U+2764`> is the unicode notation for a (heavy) heart while `<U+2606>` is unicode for a white star. You can type the unicode value into any web search engine to see the actual symbol.


Suppose we wish to clean up the file by removing all the unicode values. For each "<", we need to find the matching ">". The pattern "[^<]*" is used to match any character except "<" zero or more times. We replace this pattern with a space.


```{r}
out7 <- str_replace_all(love, "<[^<]*>", " ")
out7[c(12, 195)]
```

To remove the *, (,  ) and `, we will need to enclose them in brackets since these are meta-characters.


```{r}
out8 <- str_replace_all(out7, "[*()`]", " ")
head(out8)
out7[c(12, 195)]
```


We can also remove extra whitespace in the text. We will replace one or more occurrences of a space (`\\s+`) with just a single space.


```{r}
out9 <- str_replace_all(out8, "\\s+", " ")
head(out9)
out9[c(12, 195)]
```

### Own your own

1. How many tweets in `love` contain the phrase "love you"?

2. How many mentions (of specific users) are there in the `love` tweets? Mentions are prefaced by the "at" symbol, for example, `@username`.

3. The third installment of the movie franchise "Ghostbusters" was released on July 15, 2016. 
Import (via 'scan') the file **Ghostbusters.txt88 that has 5000 tweets downloaded from Twitter on July 18, 2016, based on a search of the word "ghostbusters". 


   * Clean up the data set by removing the unicode values.

   * How many mentions are there of the word "fun"?

   * How many of the 5000 tweets are retweets?

   * How many tweets mention the actress Kate McKinnon? Take into account capitalization of her name (including "MCK" and "McK").

   * How many occurrences are there of the word "love"? Take into account spellings such as "LOOOVED".


### Resources

* To download tweets from Twitter, you need to have a Twitter account and then sign into the developers page. Details on the authentication procedure can be found at 
[http://thinktostart.com/twitter-authentification-with-r/](http://thinktostart.com/twitter-authentification-with-r/)