---
title: "Tornadoes in Texas 2012-2016"
output: word_document
editor_options: 
  chunk_output_type: console
---
Tornadoes are very destructive weather events that can cause deaths and much property damage. Their severity are measured on the enhanced Fujita scale:

F Scale | Speed 
--------|-------------------
EF0     | 65-85 mph (light damage)
EF1     | 86-110 mph (moderate damage)
EF2     | 111-135 mph (considerable damage)
EF3     | 136-165 mph (severe damage)
EF4     | 166-200 mph (devastating damage)
EF5     | > 200 mph (incredible damage)


In which months do tornadoes tend to occur? What time? Is there a pattern to where the strongest tornadoes occur?


We will investigate tornadoes in the state of Texas from March 3, 2012 through March 3, 2016. The data, in the file **tornadoes12-16.txt**, were obtained from the Storm Events Database at the National Centers for Environmental Information [www.ncdc.noaa.gov](www.ncdc.noaa.gov)


We will need the packages `stringr`, `lubridate`, `ggplot2`, `dplyr` and `plyr` for this lab:


```{r, message = FALSE}
for (package in c("stringr", "lubridate", "ggplot2", "plyr", "dplyr", "ggformula")){
  library(package, character.only = TRUE)
}

```


The values in the variable `BeginTime` (the time the tornado began) in the text  file are quoted because some of the times have leading zeros (for example, "0305" for 3:05 am). However, even with the double quotes, R infers that this column is numeric and strips off the leading zeros. We will use the `colClasses` argument to `read.table` to force R to read this variable as a character.


```{r}
tornadoes <- read.table("TxTornadoes12-16.txt",   
        colClasses=c(Fscale = "factor", Source = "factor", 
        BeginTime="character"), stringsAsFactors = FALSE, 
        header = TRUE, sep = ",")
```


The date (`BeginDate`) and time (`BeginTime`) of the tornadoes are given as separate variables so we will combine them into one variable. In addition, the time variable will also need to be converted into the hh:mm:ss format.


The `BeginTime` variable has the hour and minute information of when the tornado began, but without the ":" separator. We will use the `str_sub` command from the `stringr` package to extract the hours and minutes portion from this variable and then use the base R command `paste` to glue them back together with the ":" separator.


```{r}
head(tornadoes$BeginTime)
temp1 <- str_sub(tornadoes$BeginTime, 1, 2)  
head(temp1)

temp2 <- str_sub(tornadoes$BeginTime, 3, 4)  
head(temp2)

temp3 <- paste(temp1, ":", temp2, ":", "00", sep = "")
head(temp3)

temp4 <- paste(tornadoes$BeginDate, temp3, sep = " ")
head(temp4)

temp5 <- mdy_hms(temp4)
head(temp5)
tornadoes$DateTime <- temp5
```

**Remark** All the times in the data set were given in Central Standard Time (CST). For the purposes of this lab, ignore the default UTC designation in the output.


When do most tornados occur? Do they tend to occur in the same months from year to year? To answer this, we can use base R and `ggplot2` commands which can handling date objects.


```{r}

tornadoes$Month <- month(tornadoes$DateTime, label = TRUE)
gf_bar( ~ Month, data = tornadoes)

table(tornadoes$Month)

tornadoes$Year <- year(tornadoes$DateTime)
gf_bar(~ as.factor(Year), data = tornadoes, fill = ~Month, position = "fill")

```

Thus, we see that 220 tornados occurred in May and only 1 occurred in January. We can also see that the proportion of a year's tornados that occur in a given month is not consistent year to year. For example, in this data set, 2015 had a higher fraction of December tornadoes than in other years. 


**Remarks** 

* Reminder: The data are for tornadoes that occurred between March 1, 2012 and March 1, 2016.

* To compute proportions instead of counts, use the `prop.table` command. For example

```{r}
prop.table(table(tornadoes$Year, tornadoes$Month), 1)
```

What time did these tornadoes occur? 


```{r}
TorTime <- hour(tornadoes$DateTime)
head(TorTime)

tornadoes$TimeOfDay <- cut(TorTime, breaks=c(0,6,12,18, 24), include.lowest = TRUE)

gf_bar( ~ TimeOfDay, data = tornadoes)
table(tornadoes$TimeOfDay)
```


Thus, we can see that most tornadoes occur between noon and midnight. 


How strong were the tornadoes that hit Texas? Is there any relationship between strength of tornado and time of occurrence?


```{r}
ggplot(tornadoes, aes(x = Fscale)) + geom_bar()

prop.table(table(tornadoes$Fscale, tornadoes$TimeOfDay))
```

We see that most tornadoes in Texas were EF0.
About 43.4% of EF1 tornadoes occurred between noon and 6pm, while 29.2% of them occurred between 6pm and midnight. 


Other common operations work on dates also. For instance, suppose we want to know what fraction of tornadoes occurred after June 30, 2014. 


```{r}
AfterJune30 <- tornadoes$DateTime > mdy_hms("06-30-2014 23:59:59")
head(AfterJune30)
mean(AfterJune30)
```

About 56.9% of tornadoes in this data set occurred after June 30, 2014.



### Mapping the tornadoes

We can visualize the occurrences of tornadoes on a map. First, some preparation.


The enhanced Fujita scale variable, `Fscale` is a factor variable but we need a numeric version for one of the `ggplot2` commands. We will use the `plyr` package to map the levels of this variable to the numeric values 0, 1, 2, 3, 4.


```{r}
tornadoes$F2 <- as.numeric(mapvalues(tornadoes$Fscale, from = levels(tornadoes$Fscale), to = 0:4))
```


Now, we will use the command `map_data` from `ggplot2` to get the outline for the state of Texas. We will store this portion of the plot in the R object `p`.


```{r}
Txmap <- map_data("state", region="Texas")

 gf_polygon(lat ~ long, data = Txmap, fill="white")

```


Now, we will plot the latitude and longitude of where each tornado began. Since we want to add a layer to the Texas map, we will using the piping operator.


```{r}
 gf_polygon(lat ~ long, data = Txmap, fill="white") %>%
  gf_point(BeginLat ~ BeginLon, data = tornadoes, size = ~ F2,   color=~Fscale) %>%
  gf_labs(x = "longitude", y = "latitude", size = "Fscale")
```

To modify the legend:

```{r}
last_plot()  +
  scale_size_continuous(labels=c("EF0", "EF1", "EF2", "EF3", "EF4"))

```

We can also see if there is a pattern to when the tornadoes occurred:

```{r}
 gf_polygon(lat ~ long, data = Txmap, fill="white")%>% 
  gf_point(BeginLat ~ BeginLon, data = tornadoes,  color= ~Month)

```


**Question 1** Is there any geographic pattern to who (storm chasers, the public, etc.) reports the tornado (variable `Source`)?


We saw earlier that most tornadoes were either EF0 or EF1 on the enhanced Fujita scale. Let's map where the more severe tornadoes occurred.


```{r}
severeTorn <- filter(tornadoes, Fscale=="EF2" | Fscale=="EF3"| Fscale=="EF4")   

 gf_polygon(lat ~ long, data = Txmap, fill="white") %>% 
  gf_point(BeginLat ~ BeginLon, data = tornadoes,  size = .5)  %>%
  gf_text(BeginLat ~ BeginLon, data = tornadoes, label = ~BeginLocation, size = 1.5, vjust = 1)  %>%
  gf_labs(x="longitude", y = "latitude", title = "Severe tornadoes")
```


The `vjust` argument to `geom_text` shifts the label up or down from the point.


**Question 2** Where were the tornadoes that were spotted by storm chasers?


###On Your Own
The file **FlLightning10-15.txt** contains data on lightning strikes in Florida between 2010 and 2015. A lightning strike was recorded if there was any property or crop damage, or injuries or deaths (either direct or indirect). Analyze this data set following the steps in the analysis of the Texas tornadoes. This file is comma delimited and the variable `Time` should be imported as character and `Source` as factor.
